\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-11-14 chapters/covert_align.tex
Started processing comments MDV

\chapter{Compensating \acs{erp} jitter for gaze-independence}
\label{sec:covert-align}
\emph{This chapter was published as part of~\textcite{VanDenKerchove2024}.}

@@

\textcite{Arico2014} observed higher variability in single-trial P3 peak
latencies relative to stimulus onset during covert \ac{vsa} compared to overt
\ac{vsa}.
This latency variability contributes to reduced covert \ac{vsa} decoding performance.
While they proposed an analysis and performance prediction method, they did not provide a
decoding solution.
They suggested that compensating for latency jitter could enhance covert \ac{vsa}
decoding, but did not verify this hypothesis directly.
Additionally, \cite{Hardiansyah2020} developed a classifier for
covert \ac{vsa} \acp{erp}, exploiting single-trial latency features in combination with
amplitude features for classification with a support vector machine.
They demonstrated the positive influence of single-trial \ac{erp} component latency
features on covert \ac{vsa} inference, yet did not attempt to correct the amplitude
features for these latencies.

@@

\subsubsection{CVSA-ERP dataset}
\label{sec:covert-align/stimulation}

We recorded a dataset to validate our approach.
The Covert Visuospatial Attention \ac{erp} (CVSA-ERP) dataset
consists of 15 participants, mean age $26.34\pm3.04$ years.
This study was approved by the Ethics Commission of University Hospital Leuven
(S62547).
Each subject performed different \ac{vsa} conditions (overt, covert, and split),
illustrated in Figure~\ref{fig:interface} (top row).
Using a hexagonal layout interface, similar to the visual Hex-o-Spell proposed
by \cite{Treder2010}, we presented six flashing targets (without letters or
symbols) to the participant while the \ac{eeg}, electrooculogram (EOG), and the
participant's eye gaze using eye tracking were recorded.
The \ac{vsa} conditions described in the first row of Figure~\ref{fig:interface} are
considered.
\begin{figure*}
  \bigskip
	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:interface}
		\includegraphics[width=\linewidth]{figures/covert_align/figure3a.pdf}
	\end{subfigure}

	\bigskip
	\bigskip
	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:erps}
		\includegraphics[width=\linewidth]{figures/covert_align/figure3b.pdf}
	\end{subfigure}%

  \caption[Stimulation interfaces and evoked \acsp{erp}.]{
		\subref{fig:interface} Interfaces and visuospatial attention (\ac{vsa})
    conditions in the CVSA-ERP and BNCI2014-009 datasets.
		In the CVSA-ERP oddball \ac{bci} interface, screen targets are intensified one after
    the other in pseudorandom order while the participant can either pay overt,
    covert, or split \ac{vsa} to the cued target.
		In the BNCI2014-009 overt \ac{vsa} interface, entire rows and columns are
		intensified at once. In its covert counterpart,
		groups of 6 letters are intensified one after the other, partly relying on
    feature attention.
		\subref{fig:erps} Contrast between target (color) and non-target, and
    distractor (gray) and non-target grand average event-related potentials per \ac{vsa} condition and dataset.
    Overt \ac{vsa} yields a strong modulation of the N1 component in both datasets;
    the P3 amplitude decreases with the degree of split \ac{vsa}.
    In split \ac{vsa}, N1 and P2 are more prominently evoked by the distractor,
    while the P3 is evoked by the target.
	}
\end{figure*}

In contrast to the protocol proposed by \cite{Frenzel2011}, split \ac{vsa} was
performed by instructing the participant to attend to the intensifications of the
cued target and ignore the intensifications of the distractor target.
Since we assume there will be an effect depending on the distance between
the attended target and the distractor, we discern three split \ac{vsa} sub-conditions:
the distractor is either clockwise or counterclockwise directly next to the
attended target ($d=1$), there is one other target between the attended target and
the distractor ($d=2$), or the distractor is opposite the intended target
($d=3$).

@@

Participants signed an informed consent form and were seated at a distance of
60 cm from a CRT-emulating monitor (VPixx
Technologies, Canada) operating at a refresh rate of 120Hz, displaying 6
circular white targets with a diameter of 4.15° visual angle and laid out on a hexagon
with a radius of 12.28° of visual angle centered on the midpoint of the screen,
conforming to the interface proposed by \textcite{Treder2010}
(\cref{fig:covert-align/interface/interface}).
A hexagonal layout interface with an empty center and a low number of targets
counteracts target crowding, and as long as the subject’s gaze is within the hexagon of
targets, no other target can be between the subject’s gaze and a covertly
attended target.
Targets are full-contrast white and were intensified by scaling them to a
larger size (5.60° of visual angle,
(\cref{fig:covert-align/interface/intensification}) instead of changing the contrast to avoid Troxler-fading\footnote{The optical illusion of disappearing unchanging stimuli
experienced when visually fixating~\cite{Troxler1804}.}~\cite{Treder2010} in the
peripheral visual field.
Stimuli were presented using PsychoPy (version 2023.1.3)~\cite{Peirce2019}.

\begin{figure}
  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/covert_align/interface_1.tikz.tex}
    \caption[The stimulation interface.]{The stimulation interface based on the visual Hex-o-Spell
    \acs{bci}~\cite{Treder2010}.}
    \label{fig:covert-align/interface/interface}
  \end{subfigure}\hfill%
  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/covert_align/interface_2.tikz.tex}
    \caption[Intensification.]{When a target is intensified, it is enlarged for
    a very short time.}
    \label{fig:covert-align/interface/intensification}
  \end{subfigure}

  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/covert_align/interface_3.tikz.tex}
    \caption[Gaze fixated on a target.]{In the overt and split \ac{vsa}
    settings, the
    participant is cued to fixate on one of the targets.}
    \label{fig:covert-align/interface/overt}
  \end{subfigure}\hfill%
  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/covert_align/interface_4.tikz.tex}
    \caption[Gaze fixated centrally.]{In the covert \ac{vsa} setting, the participant is
    cued to fixate at the center of the interface.}
    \label{fig:covert-align/interface/covert}
  \end{subfigure}
\end{figure}

@@

\subsubsection{BNCI2014-009 dataset}
The BNCI2014-009\footnote{\url{https://bnci-horizon-2020.eu/database/data-sets}}
dataset~\cite{Aloise2012a} was used in the analysis performed
by \cite{Arico2014}.
It contains data from 10 subjects (median age $24.5\pm1.9$
years) who performed two spelling tasks illustrated in the second row of
\cref{fig:interface}: using the P3 Matrix speller
interface to exploit overt \ac{vsa}, and the GeoSpell covert \ac{vsa} interface.
To use the GeoSpell interface, the participant gazes at the fixation point at the
center of the screen, while groups of characters flash simultaneously in a
circular layout around the fixation point.
The user directs their visuospatial attention to the location where the intended letter is expected
to appear, and when it does, a P3 \ac{erp} component is expected to be evoked.
This results in a specific setting where both visuospatial attention and
feature attention (the attended letter) are exploited.
For a detailed description of the paradigm and dataset, we refer
to \cite{Aloise2012a}.

@@

\subsubsection{Decoders}
\label{sec:covert-align/methods/decoders}
We compared \ac{wcble} and \ac{cble} as described in
\cref{sec:wcble/methods}, with their base classifier \ac{tlda} and with
Riemannian Geometry approaches that rely on spatial covariance as features.
Together with \ac{tlda}, Riemannian Geometry generally achieves state-of-the-art decoding
performance~\cite{Lotte2018}.
We implemented two Riemannian Geometry pipelines.
The first estimates shrunk covariances from the \acp{erp} filtered with 6 XDAWN
filters, projects these covariances to a tangent space, and classifies the
result using $L_2$-regularized logistic regression
(XDAWNCov-TS-LR)~\cite{Cecotti2017}.
Secondly, we adopted the pipeline from \cite{Aydarkhanov2020}, as their work shows
favorable performance in the presence of single-trial \ac{erp} latency jitter.
Shrunk spatial covariance matrices were estimated from epochs that were
augmented by concatenating the average target and average non-target \ac{erp} as
extra channels, projected to tangent space, and classified using
$L_2$-regularized logistic regression (ERPCov-TS-LR).

@@

\subsection{\Acs{bci} decoding performance}
\label{sec:block_accuracy}

@@

\Cref{fig:single_trial_roc_auc_diff} shows a comparison of \ac{rocauc}
for all pairs of \ac{tlda}, \ac{cble},
and \ac{wcble} for single-trial classification to investigate the contributions of
\ac{cble} and \ac{wcble} relative to their first-stage classifier \ac{tlda}.
For this evaluation, epochs were rejected when the peak-to-peak amplitude
exceeded 800$\mu V$, and for the CVSA-ERP dataset, if the user's gaze differed by more than
10 degrees of visual angle from the fixation crosshair.
\begin{figure}
  \bigskip
	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:single_trial_roc_auc_diff}
		\includegraphics[width=\linewidth]{figures/covert_align/figure4a.pdf}
	\end{subfigure}

@@

	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:block_evaluation}
		\includegraphics[width=\linewidth]{figures/covert_align/figure4b.pdf}
	\end{subfigure}
  \caption[Decoder performance.]{%
		\subref{fig:single_trial_roc_auc_diff} Difference in cross-validated
    single-trial classification receiver operating characteristic curve ($\Delta$ROC-AUC)
    between Classifier-based Latency Estimation (CBLE), Woody
    CBLE (WCBLE),	and their first-stage classifier (tLDA).
    95\% confidence intervals were determined using $k=1000$ bootstrapping.
    Our proposed WCBLE decoder outperforms tLDA and CBLE for covert and split
		($d=3$)	visuospatial attention (VSA). CBLE scores on par with tLDA.
		\subref{fig:block_evaluation} Cross-validated target selection accuracy for
    all decoders plotted as a function of the number of test repetitions in
    different VSA conditions. Significance was determined using one-sided
    (WCBLE $>$ other) Wilcoxon rank-sum tests using False Discovery Rate correction
    ($*= p<0.05$, $**=p<0.01$, $***=p<0.001$). WCBLE generally achieves the highest
    covert VSA target selection accuracy.}
\end{figure}
Wilcoxon signed-rank tests controlled for multiple comparisons by Benjamini
and Hochberg's False Discovery Rate procedure (FDR) revealed that
for the BNCI2014-009 dataset, \ac{wcble} significantly outperformed \ac{tlda}
($\Delta\mathrm{ROC-AUC} = 0.019$, $p=0.004$) and \ac{cble}
($\Delta\mathrm{ROC-AUC} = 0.016$, $p=0.036$) for covert \ac{vsa},
but was significantly outperformed by \ac{tlda} in overt \ac{vsa} decoding
($\Delta\mathrm{ROC-AUC}=-0.004$, $p=0.040$).
For the CVSA-ERP dataset, \ac{wcble}
also achieved significantly better covert \ac{vsa} performance than \ac{tlda} ($\Delta\mathrm{ROC-AUC}
	= 0.023$, $p=0.041$) and \ac{cble} ($\Delta\mathrm{ROC-AUC}
	= 0.036$, $p=0.024$).
We found no significant difference in \ac{wcble} performance over \ac{tlda} in the split
\ac{vsa} conditions in the CVSA-ERP dataset, but results show a clear trend of
increased \ac{wcble} performance over \ac{tlda} and \ac{cble} as $d$ increases.
\ac{cble} failed to significantly outperform its first-stage classifier \ac{tlda} in all
evaluated \ac{vsa} conditions.
\Cref{tab:covert-align/results/single-trial} reports all single-trial classification scores for all considered models, datasets,
and conditions.

@@

\Cref{fig:block_evaluation} shows the cross-validated \ac{bci} selection
accuracy on the BNCI2014-009 and the CVSA-ERP datasets for all investigated
decoders.
Accuracy was determined by, for each block, selecting the character with the
highest (stage-two if applicable) classifier score and comparing it to the cued target.
Significance was calculated using one-sided Wilcoxon rank-sum tests ($p=0.05$)
corrected for FDR over decoders.
For this evaluation, no epochs were rejected to keep the trial-based structure
of \ac{bci} operation intact.
For all datasets and \ac{vsa} conditions, \ac{cble} scores approximately on par with \ac{tlda}.
Yet, \ac{wcble} yields an improved decoding accuracy for covert \ac{vsa} in both
datasets, which is greatest for smaller numbers of repetitions and
decreases as the number of repetitions increases.
This covert \ac{vsa} accuracy increase over \ac{tlda} is significant in thdeep q learning tutoriale BNCI2014-009 dataset
for 1 and 3 repetitions, and in CVSA-ERP for 1-5 and 10 repetitions.
Furthermore, while we reported a relative decrease in single-trial \ac{rocauc} for
\ac{wcble} in overt \ac{vsa}, this does not seem to result in a consistent decrease in target
selection accuracy.
No significant increase of \ac{wcble} over other methods was found in split \ac{vsa}.
While Riemannian methods are significantly outperformed by \ac{tlda}, \ac{cble}, and \ac{wcble}
in the BNCI2014-009 dataset, they perform approximately on par with \ac{tlda} and
\ac{wcble} in CVSA-ERP.

@@

\subsection{Gaze-independence through cross-condition transfer}
\label{sec:cross_results}
To further back our claim of gaze-independence in the case where eye motor
control cannot be assumed, we evaluate our proposed decoder in a transfer
learning setting between \ac{vsa} conditions.
While performing overt attention requires gaze redirection for each target
selection, performing covert or split \ac{vsa} continuously still requires
sustained gaze fixation, which might be impaired in those who
could benefit from such an application.
Studying the transfer between conditions simulates what happens
when the user performs different \ac{vsa} conditions throughout the
experimental session.
Furthermore, if our decoder performs well in transfer-learning settings,
it must capture some information about the \ac{erp} responses that is independent of
the \ac{vsa} condition, and hence does not depend on gaze redirection to perform
these conditions.
We introduce an additional setting of interest here, namely a combination of \ac{vsa}
conditions, which represents those cases where the user cannot
redirect their gaze and hence can be in any one of the \ac{vsa} conditions depending on the
target they attend to.
For BNCI2014-009, this is implemented as an equal mix of overt and
covert \ac{vsa}; for CVSA-ERP, the combined condition represents an equal mix of overt,
covert, and split \ac{vsa}, disregarding parameter $d$.

Figure~\ref{fig:covert_cross_eval} and Figure~\ref{fig:aloise2012_cross_eval} show
the pairwise differences in area under the ROC curve ($\Delta$ROC-AUC)
between the investigated decoders.
In this evaluation, bad epochs were rejected as in \cref{sec:block_accuracy}.
\begin{figure*}
  \bigskip
	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:covert_cross_eval}
		\includegraphics[width=\linewidth]{figures/covert_align/figure5a.pdf}
	\end{subfigure}

@@

	\begin{subfigure}[c]{.48\linewidth}
		\caption{}
		\label{fig:aloise2012_cross_eval}
		\includegraphics[width=\linewidth]{figures/covert_align/figure5b.pdf}
	\end{subfigure}\hfill%
	\begin{subfigure}[c]{.48\linewidth}
		\caption{}
		\label{fig:jitter}
		\includegraphics[width=\linewidth]{figures/covert_align/figure5c.pdf}
	\end{subfigure}

  \caption[Cross-condition classifier performance and estimated jitter.]{
		\subref{fig:covert_cross_eval},\subref{fig:aloise2012_cross_eval}
    Difference in cross-validated area under the receiver
		operating characteristic curves between Classifier-based Latency Estimation
    (CBLE) and Woody CBLE (WCBLE) across conditions for the CVSA-ERP and
    BNCI2014-009 datasets, respectively.
    A decoder is each	time trained on a visuospatial attention (VSA) condition
    and tested on all VSA conditions.
    WCBLE yields an improvement in most
		non-overt \ac{vsa} settings, indicating it is more invariant to eye gaze than
    CBLE and \ac{tlda}.
		\subref{fig:jitter} Jitter characterized as the interquartile range (IQR)
		of target epochs for different \ac{vsa} conditions. Overt \ac{vsa} exhibits lower
    jitter than other conditions.
    Significance of differences
    was determined with two-sided Wilcoxon rank-sum tests with False Discovery
    Rate correction on per-subject jitter ($*= p<0.05$, $**=p<0.01$,
    $***=p<0.001$, $****=p<0.0001$).
	}
\end{figure*}
When comparing \ac{cble} and \ac{tlda}, we do not observe large differences in any of the
evaluated settings, similar to the within-subject conditions.
On the contrary, when considering the comparisons between \ac{wcble} and \ac{tlda}, we
see that performance is on par or greater using \ac{wcble} for most conditions,
except for within-overt \ac{vsa} decoding.

>>>>

\chapter{Compensating \acs{erp} jitter for gaze-independence}%
\label{sec:covert-align}
\emph{This chapter was published as part of~\textcite{VanDenKerchove2024}.}

@@

\textcite{Arico2014} observed higher variability in single-trial P3 peak
latencies relative to stimulus onset during covert \ac{vsa} compared to overt
\ac{vsa}.
This latency variability contributes to reduced covert \ac{vsa} decoding performance.
While they proposed an analysis and performance prediction method, they did not provide a
decoding solution.
They suggested that compensating for latency jitter could enhance covert \ac{vsa}
decoding, but did not verify this hypothesis directly.
Additionally, \textcite{Hardiansyah2020} developed a classifier for
covert \ac{vsa} \acp{erp}, exploiting single-trial latency features in combination with
amplitude features for classification with a support vector machine.
They demonstrated the positive influence of single-trial \ac{erp} component latency
features on covert \ac{vsa} inference, yet did not attempt to correct the amplitude
features for these latencies.

@@

\subsubsection{CVSA-ERP dataset}%
\label{sec:covert-align/stimulation}

We recorded a dataset to validate our approach.
The Covert Visuospatial Attention \ac{erp} (CVSA-ERP) dataset
consists of 15 participants, mean age $26.34\pm3.04$ years.
This study was approved by the Ethics Commission of University Hospital Leuven
(S62547).
Each subject performed different \ac{vsa} conditions (overt, covert, and split),
illustrated in Figure~\ref{fig:interface} (top row).
Using a hexagonal layout interface, similar to the visual Hex-o-Spell proposed
by \textcite{Treder2010}, we presented six flashing targets (without letters or
symbols) to the participant while the \ac{eeg}, electrooculogram (EOG), and the
participant's eye gaze using eye tracking were recorded.
The \ac{vsa} conditions described in the first row of Figure~\ref{fig:interface} are
considered.
\begin{figure*}
  \bigskip
	\begin{subfigure}{\linewidth}
		\caption{}%
		\label{fig:interface}
		\includegraphics[width=\linewidth]{figures/covert_align/figure3a.pdf}
	\end{subfigure}

	\bigskip
	\bigskip
	\begin{subfigure}{\linewidth}
		\caption{}%
		\label{fig:erps}
		\includegraphics[width=\linewidth]{figures/covert_align/figure3b.pdf}
	\end{subfigure}%

  \caption[Stimulation interfaces and evoked \acsp{erp}.]{%
		\subref{fig:interface} Interfaces and visuospatial attention (\ac{vsa})
    conditions in the CVSA-ERP and BNCI2014-009 datasets.
		In the CVSA-ERP oddball \ac{bci} interface, screen targets are intensified one after
    the other in pseudorandom order while the participant can either pay overt,
    covert, or split \ac{vsa} to the cued target.
		In the BNCI2014-009 overt \ac{vsa} interface, entire rows and columns are
		intensified at once. In its covert counterpart,
		groups of 6 letters are intensified one after the other, partly relying on
    feature attention.
		\subref{fig:erps} Contrast between target (color) and non-target, and
    distractor (gray) and non-target grand average event-related potentials per \ac{vsa} condition and dataset.
    Overt \ac{vsa} yields a strong modulation of the N1 component in both datasets;
    the P3 amplitude decreases with the degree of split \ac{vsa}.
    In split \ac{vsa}, N1 and P2 are more prominently evoked by the distractor,
    while the P3 is evoked by the target.
	}
\end{figure*}

In contrast to the protocol proposed by \textcite{Frenzel2011}, split \ac{vsa} was
performed by instructing the participant to attend to the intensifications of the
cued target and ignore the intensifications of the distractor target.
Since we assume there will be an effect depending on the distance between
the attended target and the distractor, we discern three split \ac{vsa} sub-conditions:
the distractor is either clockwise or counterclockwise directly next to the
attended target ($d=1$), there is one other target between the attended target and
the distractor ($d=2$), or the distractor is opposite the intended target
($d=3$).

@@

Participants signed an informed consent form and were seated at a distance of
60 cm from a CRT-emulating monitor (VPixx
Technologies, Canada) operating at a refresh rate of 120Hz, displaying 6
circular white targets with a diameter of 4.15° visual angle and laid out on a hexagon
with a radius of 12.28° of visual angle centered on the midpoint of the screen,
conforming to the interface proposed by \textcite{Treder2010}
(\cref{fig:covert-align/interface/interface}).
A hexagonal layout interface with an empty center and a low number of targets
counteracts target crowding, and as long as the subject’s gaze is within the hexagon of
targets, no other target can be between the subject’s gaze and a covertly
attended target.
Targets are full-contrast white and were intensified by scaling them to a
larger size (5.60° of visual angle, see \cref{fig:covert-align/interface/intensification})
instead of changing the contrast to avoid Troxler-fading\footnote{The optical illusion of disappearing unchanging stimuli
experienced when visually fixating~\cite{Troxler1804}.}~\cite{Treder2010} in the
peripheral visual field.
Stimuli were presented using PsychoPy (version 2023.1.3)~\cite{Peirce2019}.

\begin{figure}
  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/covert_align/interface_1.tikz.tex}
    \caption[The stimulation interface.]{The stimulation interface based on the visual Hex-o-Spell
    \acs{bci}~\cite{Treder2010}.}%
    \label{fig:covert-align/interface/interface}
  \end{subfigure}\hfill%
  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/covert_align/interface_2.tikz.tex}
    \caption[Intensification.]{When a target is intensified, it is enlarged for
    a very short time.}%
    \label{fig:covert-align/interface/intensification}
  \end{subfigure}

  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/covert_align/interface_3.tikz.tex}
    \caption[Gaze fixated on a target.]{In the overt and split \ac{vsa}
    settings, the
    participant is cued to fixate on one of the targets.}%
    \label{fig:covert-align/interface/overt}
  \end{subfigure}\hfill%
  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/covert_align/interface_4.tikz.tex}
    \caption[Gaze fixated centrally.]{In the covert \ac{vsa} setting, the participant is
    cued to fixate at the center of the interface.}%
    \label{fig:covert-align/interface/covert}
  \end{subfigure}
\end{figure}

@@

\subsubsection{BNCI2014-009 dataset}
The BNCI2014-009\footnote{\url{https://bnci-horizon-2020.eu/database/data-sets}}
dataset~\cite{Aloise2012a} was used in the analysis performed
by \textcite{Arico2014}.
It contains data from 10 subjects (median age $24.5\pm1.9$
years) who performed two spelling tasks illustrated in the second row of
\cref{fig:interface}: using the P3 Matrix speller
interface to exploit overt \ac{vsa}, and the GeoSpell covert \ac{vsa} interface.
To use the GeoSpell interface, the participant gazes at the fixation point at the
center of the screen, while groups of characters flash simultaneously in a
circular layout around the fixation point.
The user directs their visuospatial attention to the location where the intended letter is expected
to appear, and when it does, a P3 \ac{erp} component is expected to be evoked.
This results in a specific setting where both visuospatial attention and
feature attention (the attended letter) are exploited.
For a detailed description of the paradigm and dataset, we refer
to \textcite{Aloise2012a}.

@@

\subsubsection{Decoders}%
\label{sec:covert-align/methods/decoders}
We compared \ac{wcble} and \ac{cble} as described in
\cref{sec:wcble/methods}, with their base classifier \ac{tlda} and with
Riemannian Geometry approaches that rely on spatial covariance as features.
Together with \ac{tlda}, Riemannian Geometry generally achieves state-of-the-art decoding
performance~\cite{Lotte2018}.
We implemented two Riemannian Geometry pipelines.
The first estimates shrunk covariances from the \acp{erp} filtered with 6 XDAWN
filters, projects these covariances to a tangent space, and classifies the
result using $L_2$-regularized logistic regression
(XDAWNCov-TS-LR)~\cite{Cecotti2017}.
Secondly, we adopted the pipeline from \textcite{Aydarkhanov2020}, as their work shows
favorable performance in the presence of single-trial \ac{erp} latency jitter.
Shrunk spatial covariance matrices were estimated from epochs that were
augmented by concatenating the average target and average non-target \ac{erp} as
extra channels, projected to tangent space, and classified using
$L_2$-regularized logistic regression (ERPCov-TS-LR).

@@

\subsection{\Acs{bci} decoding performance}%
\label{sec:covert-align/results/bloc-acc}

@@

\Cref{fig:covert-align/single-trial-roc-auc-diff} shows a comparison of \ac{rocauc}
for all pairs of \ac{tlda}, \ac{cble},
and \ac{wcble} for single-trial classification to investigate the contributions of
\ac{cble} and \ac{wcble} relative to their first-stage classifier \ac{tlda}.
For this evaluation, epochs were rejected when the peak-to-peak amplitude
exceeded 800$\mu V$, and for the CVSA-ERP dataset, if the user's gaze differed by more than
10 degrees of visual angle from the fixation crosshair.
\begin{figure}
  \bigskip
	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:covert-align/single-trial-roc-auc-diff}%
		\includegraphics[width=\linewidth]{figures/covert_align/figure4a.pdf}
	\end{subfigure}

@@

	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:covert-align/block-eval}%
		\includegraphics[width=\linewidth]{figures/covert_align/figure4b.pdf}
	\end{subfigure}
  \caption[Decoder performance.]{%
		\subref{fig:covert-align/single-trial-roc-auc-diff} Difference in cross-validated
    single-trial classification \acl{rocauc} ($\Delta$ROC-AUC)
    between \ac{cble}, \ac{wcble}, and their first-stage classifier \ac{tlda}.
    95\% confidence intervals were determined using $k=1000$ bootstrapping.
    Our proposed \ac{wcble} decoder outperforms \ac{tlda} and \ac{cble} for covert and split
    ($d=3$)	\ac{vsa}. \ac{cble} scores on par with \ac{tlda}.
		\subref{fig:covert-align/block-eval} Cross-validated target selection accuracy for
    all decoders plotted as a function of the number of test repetitions in
    different VSA conditions. Significance was determined using one-sided
    (\ac{wcble} $>$ other) Wilcoxon rank-sum tests using False Discovery Rate correction
    ($*= p<0.05$, $**=p<0.01$, $***=p<0.001$). \ac{wcble} generally achieves the highest
    covert \ac{vsa} target selection accuracy.}
\end{figure}
Wilcoxon signed-rank tests controlled for multiple comparisons by Benjamini
and Hochberg's False Discovery Rate procedure (FDR) revealed that
for the BNCI2014-009 dataset, \ac{wcble} significantly outperformed \ac{tlda}
($\Delta\mathrm{ROC-AUC} = 0.019$, $p=0.004$) and \ac{cble}
($\Delta\mathrm{ROC-AUC} = 0.016$, $p=0.036$) for covert \ac{vsa},
but was significantly outperformed by \ac{tlda} in overt \ac{vsa} decoding
($\Delta\mathrm{ROC-AUC}=-0.004$, $p=0.040$).
For the CVSA-ERP dataset, \ac{wcble}
also achieved significantly better covert \ac{vsa} performance than \ac{tlda} ($\Delta\mathrm{ROC-AUC}
	= 0.023$, $p=0.041$) and \ac{cble} ($\Delta\mathrm{ROC-AUC}
	= 0.036$, $p=0.024$).
We found no significant difference in \ac{wcble} performance over \ac{tlda} in the split
\ac{vsa} conditions in the CVSA-ERP dataset, but results show a clear trend of
increased \ac{wcble} performance over \ac{tlda} and \ac{cble} as $d$ increases.
\ac{cble} failed to significantly outperform its first-stage classifier \ac{tlda} in all
evaluated \ac{vsa} conditions.
\Cref{tab:covert-align/results/single-trial} reports all single-trial classification scores for all considered models, datasets,
and conditions.

@@

\Cref{fig:covert-align/block-eval} shows the cross-validated \ac{bci} selection
accuracy on the BNCI2014-009 and the CVSA-ERP datasets for all investigated
decoders.
Accuracy was determined by, for each block, selecting the character with the
highest (stage-two if applicable) classifier score and comparing it to the cued target.
Significance was calculated using one-sided Wilcoxon rank-sum tests ($p=0.05$)
corrected for FDR over decoders.
For this evaluation, no epochs were rejected to keep the trial-based structure
of \ac{bci} operation intact.
For all datasets and \ac{vsa} conditions, \ac{cble} scores approximately on par with \ac{tlda}.
Yet, \ac{wcble} yields an improved decoding accuracy for covert \ac{vsa} in both
datasets, which is greatest for smaller numbers of repetitions and
decreases as the number of repetitions increases.
This covert \ac{vsa} accuracy increase over \ac{tlda} is significant in thdeep q learning tutoriale BNCI2014-009 dataset
for 1 and 3 repetitions, and in CVSA-ERP for 1-5 and 10 repetitions.
Furthermore, while we reported a relative decrease in single-trial \ac{rocauc} for
\ac{wcble} in overt \ac{vsa}, this does not seem to result in a consistent decrease in target
selection accuracy.
No significant increase of \ac{wcble} over other methods was found in split \ac{vsa}.
While Riemannian methods are significantly outperformed by \ac{tlda}, \ac{cble}, and \ac{wcble}
in the BNCI2014-009 dataset, they perform approximately on par with \ac{tlda} and
\ac{wcble} in CVSA-ERP.

@@

\subsection{Gaze-independence through cross-condition transfer}
\label{sec:covert-align/results/cross}
To further back our claim of gaze-independence in the case where eye motor
control cannot be assumed, we evaluate our proposed decoder in a transfer
learning setting between \ac{vsa} conditions.
While performing overt attention requires gaze redirection for each target
selection, performing covert or split \ac{vsa} continuously still requires
sustained gaze fixation, which might be impaired in those who
could benefit from such an application.
Studying the transfer between conditions simulates what happens
when the user performs different \ac{vsa} conditions throughout the
experimental session.
Furthermore, if our decoder performs well in transfer-learning settings,
it must capture some information about the \ac{erp} responses that is independent of
the \ac{vsa} condition, and hence does not depend on gaze redirection to perform
these conditions.
We introduce an additional setting of interest here, namely a combination of \ac{vsa}
conditions, which represents those cases where the user cannot
redirect their gaze and hence can be in any one of the \ac{vsa} conditions depending on the
target they attend to.
For BNCI2014-009, this is implemented as an equal mix of overt and
covert \ac{vsa}; for CVSA-ERP, the combined condition represents an equal mix of overt,
covert, and split \ac{vsa}, disregarding parameter $d$.

Figure~\ref{fig:covert-align/covert-cross-eval} and Figure~\ref{fig:covert-align/aloise2012-cross-eval} show
the pairwise differences in area under the ROC curve ($\Delta$ROC-AUC)
between the investigated decoders.
In this evaluation, bad epochs were rejected as in \cref{sec:covert-align/results/bloc-acc}.
\begin{figure*}
  \bigskip
	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:covert-align/covert-cross-eval}
		\includegraphics[width=\linewidth]{figures/covert_align/figure5a.pdf}
	\end{subfigure}

@@

	\begin{subfigure}[c]{.48\linewidth}
		\caption{}
		\label{fig:covert-align/aloise2012-cross-eval}
		\includegraphics[width=\linewidth]{figures/covert_align/figure5b.pdf}
	\end{subfigure}\hfill%
	\begin{subfigure}[c]{.48\linewidth}
		\caption{}
		\label{fig:jitter}
		\includegraphics[width=\linewidth]{figures/covert_align/figure5c.pdf}
	\end{subfigure}

  \caption[Cross-condition classifier performance and estimated jitter.]{
		\subref{fig:covert-align/covert-cross-eval},\subref{fig:covert-align/aloise2012-cross-eval}
    Difference in cross-validated area under the receiver
		operating characteristic curves between Classifier-based Latency Estimation
    (CBLE) and Woody CBLE (WCBLE) across conditions for the CVSA-ERP and
    BNCI2014-009 datasets, respectively.
    A decoder is each	time trained on a visuospatial attention (VSA) condition
    and tested on all VSA conditions.
    WCBLE yields an improvement in most
		non-overt \ac{vsa} settings, indicating it is more invariant to eye gaze than
    CBLE and \ac{tlda}.
		\subref{fig:jitter} Jitter characterized as the interquartile range (IQR)
		of target epochs for different \ac{vsa} conditions. Overt \ac{vsa} exhibits lower
    jitter than other conditions.
    Significance of differences
    was determined with two-sided Wilcoxon rank-sum tests with False Discovery
    Rate correction on per-subject jitter ($*= p<0.05$, $**=p<0.01$,
    $***=p<0.001$, $****=p<0.0001$).
	}
\end{figure*}
When comparing \ac{cble} and \ac{tlda}, we do not observe large differences in any of the
evaluated settings, similar to the within-subject conditions.
On the contrary, when considering the comparisons between \ac{wcble} and \ac{tlda}, we
see that performance is on par or greater using \ac{wcble} for most conditions,
except for within-overt \ac{vsa} decoding.
